---
title: Causal Priority Effects
authors:
- name: Bradley Saul
- name: Fletcher Halliday
output: 
  pdf_document
header-includes: 
- \usepackage{mathtools}
- \usepackage{tikz}
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{xcolor}
- \usepackage{setspace}
- \newcommand*\circled[1]{\tikz[baseline=(char.base)]{
     \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}
bibliography: "priority_effects.bib"
abstract: TBD
---

\doublespacing

# Introduction

The early bird gets the worm, but the second mouse gets the cheese. Despite being poorly understood and studied, priority effects -- the order and timing of species arrival -- have ecological consequences [@fukami2015historical]. Priority effects invite counterfactual reasoning: if species A arrives first would outcomes for species B be different compared to a world where species B arrived first? Such counterfactual questions are well-suited for the potential framework of causal inference [@vanderweele2015explanation]. 

To study within-host priority effects, @halliday2017interactions conducted an experiment wherein three cohorts of sentinel, pathogen-naive tall fescue (*Festuca arundinacea*) plants where placed in a field at three different timepoints during the year. The timepoints were chosen to correspond to periods of high prevelance during the annual epidemic cycles for each of three tall fescue fungal pathogens: *Colletotrichum* (early summer), *Rhizoctonia* (mid summer), and *Puccinia* (late summer) [@halliday2017interactions]. 

In this paper, priority effects are defined as contrasts of potential outcomes, hence under assumptions that we detail these effects have a causal interpretation and may be estimated from observed data. 

## Priority Effects

Describe priority effects and importance to ecology (based on @fukami2015historical)

* would be nice to map one of the figures in @fukami2015historical to the causal graph

## Causal inference using potential outcomes

In this paper, causal research questions about priority effects are explicity defined in terms of potential outcomes. Using a potential outcome approach, causal parameters can be defined and estimators derived without reference to specific statistical models [@imai2010general].

# Discrete-Time Causal Models for Within-Host Priority Effects

Define $\mathsf{I}_{jkt} = (C_{jkt}, R_{jkt}, P_{jkt})^\intercal$ as the indicators of observed infection status by pathogens *Colletotrichum*, *Rhizoctonia*, and *Puccinia* at week $t = 1, \dots,$ after the start of the experiment on leaf $k = 1, \dots, n_{jt}$ in plant $j = 1, \dots, m$. Let $D_{jkt}$ be an indicator that a leaf is alive (1) or not (0) at $t$; $H_{j} \in \{1, 2, 3\}$ indicate plant $j$'s randomly assigned cohort; $A_{jkt}$ be a leaf's age (in weeks) on week $t$; and $L_{jkt}$ be additional covariates such as abiotic factors measured in the time up to the timepoint $t$,

Denote the random matrix of infection exposures for all $n_{jt}$ leaves in plant $j$ at time $t$ as $\mathsf{I}_{kt} = (\mathsf{I}_{j1t} , \dots, \mathsf{I}_{kn_jt})$, and let $\mathsf{I}_{j \setminus k, t}$ be the submatrix of $\mathsf{I}_{jt}$ of exposure for all leaves other than the $j$th one. Lower-case letters denote possible realizations of $\mathsf{I}_{jkt}$, $\mathsf{I}_{j \setminus k, t}$, and $\mathsf{I}_{k}$:  $\mathsf{i}_{jkt}$, $\mathsf{i}_{j \setminus k, t}$ and $\mathsf{i}_{kt}$, respectively. Define $\omega(\mathsf{i}_{t}; \rho) = \Pr_{\rho}(\mathsf{I}_{jt} = \mathsf{i}_{t})$ as the marginal probability under within-plant pressure $\rho = (\rho^c_{t-1}, \rho^r_{t-1}, \rho^p_{t-1})$ that a host plant experiences $\mathsf{i}_t$ at time $t$. 

A goal of the model is to make comparisons such as: 

* How does the risk of *Colletotrichum* infection change had the leaf not been infected by *Rhizoctonia* at the previous timepoint *versus* not having been infected by *Rhizoctonia*, holding within-plant *Rhizoctonia* intensity and within-plant *Colletotrichum* epidemic intensity fixed?
* How does the risk of *Rhizoctonia* infection change had the within-plant *Rhizoctonia* intensity been $\rho_r$ versus $\rho_r'$ and within-plant *Colletotrichum* intensity fixed and not having been infected by *Colletotrichum*?

## Causal effects (estimands)

For each $t$, the average potential outcomes are defined for each pathogen under the case that the pathogen infection statuses had been $\mathsf{i}_{t - 1} = (c_{t -1}, r_{t-1}, p_{t-1})$ and the within-host pressure had been $\rho$ in the previous week. Additionally, this estimand is conditional upon not having add been previously infected by the same pathogen and not having died.  Cohort assignment and leaf age are incorporated as effect modifiers, as cohort timing and leaf age are relevant to research questions. 

NOTES:

* Leaf death is a competing risk to infection. However, some of the pathogens may induce death. So, incorporate this competing risk, or just define (as here) population conditional upon leaf not dying
* as we may expect the effects to be different across cohorts, we define the effects to be different. But we do not define the potential outcomes as functions of the cohort since we do not think that the underlying causal processes are dependent on the timing of the cohort's placement in the field. Allows for effect modification by leaf age.

\begin{align*}
\mu_{aht}(\mathsf{i}_{t-1}, \rho) = 
\begin{pmatrix}
  \mu^c_{aht}(\mathsf{i}_{t-1}, \rho) \\
  \mu^r_{aht}(\mathsf{i}_{t-1}, \rho) \\
  \mu^p_{aht}(\mathsf{i}_{t-1}, \rho) 
\end{pmatrix} &= 
\mathrm{E} 
\left\{ 
\begin{array}{l}
  \bar{C}_{t}(\mathsf{i}_{t-1}, \rho) | A = a, H = h, C_{jk,t-1} = 0, D_{jkt} = 0 \\
  \bar{R}_{t}(\mathsf{i}_{t-1}, \rho) | A = a, H = h, R_{jk,t-1} = 0, D_{jkt} = 0 \\ 
  \bar{P}_{t}(\mathsf{i}_{t-1}, \rho) | A = a, H = h, P_{jk,t-1} = 0, D_{jkt} = 0
\end{array}
\right\} \\
&= 
\mathrm{E} \left\{
\begin{array}{l}
  \sum_{\mathsf{i}_{j \setminus k}} C_{t}(\mathsf{i}_{t-1}; \rho) \omega(\mathsf{i}_{t-1}; \rho) | A = a, H = h, C_{t-1} = 0, D_t = 0 \\
  \sum_{\mathsf{i}_{j \setminus k}} R_{t}(\mathsf{i}_{t-1}; \rho) \omega(\mathsf{i}_{t-1}; \rho) | A = a, H = h, R_{t-1} = 0, D_t = 0 \\ 
  \sum_{\mathsf{i}_{j \setminus k}} P_{t}(\mathsf{i}_{t-1}; \rho) \omega(\mathsf{i}_{t-1}; \rho) | A = a, H = h, P_{t-1} = 0, D_t = 0
\end{array}
\right\}
,
\end{align*}

where $\sum_{\mathsf{i}_{j \setminus k}} \equiv \sum_{\mathbf{r}_{j \setminus k}} \sum_{\mathbf{c}_{j \setminus k}} \sum_{\mathbf{p}_{j \setminus k}}$ and $\mathrm{E}(\cdot)$ is the expected value in the super-population of plants.

Table \ref{tab:estimands} provides a typology of estimands under consideration for this study. For example, $AL[(1, 0, 0)^{\intercal}, (0, 0, 0)^{\intercal}, \rho]$ would compare the average potential outcomes under the setting where a leaf is infected by *Colletotrichum* and nothing else in the previous week versus the setting where leaves where not infected by any pathogen, for a fixed $\rho$. In essense, this is a comparison of the probability of infection by each of three pathogens in week $t$ under two different settings of the previous week: a leaf is infected by *Colletotrichum* only versus no infections with the prevelance of infections within the plant held at $\rho$.

Under this setup, a variety of interesting causal questions can be asked. Whether and how these quantities can be estimated from data will be considered in Section XXX.

\begin{table}[H]
\caption{A typology of estimands for this study}
\label{tab:estimands}
\begin{tabular}{llll}
Label & Name of effect & Causal contrast of $\mu_{aht}(\cdot)$ & Description \\
\hline
$AL(\mathsf{i}_{t-1}, \mathsf{i}'_{t-1}, \rho)$ & Average Leaf-level &  $\cdot = (\mathsf{i}_{t-1}, \rho)$ vs $(\mathsf{i}'_{t-1}, \rho)$ & \\
$AS(\mathsf{i}_{t-1}, \rho, \rho')$ & Average Within-plant Spillover & $(\mathsf{i}_{t-1}, \rho)$ vs $(\mathsf{i}_{t-1}, \rho')$ & \\
$AT(\mathsf{i}_{t-1}, \mathsf{i}'_{t-1}, \rho, \rho')$  & Average Total & $(\mathsf{i}_{t-1}, \rho)$ vs $(\mathsf{i}'_{t-1}, \rho')$ & \\
\end{tabular}
\end{table}


```{r tikz-def, eval = TRUE, engine = "tikz", fig.cap = "Defintion for $I_t$ graph", fig.ext = 'pdf', echo = FALSE}
\usetikzlibrary{arrows, shapes, calc, shapes.geometric, fit}

\tikzset{
container/.style = {
    rectangle,
    draw,
    inner sep=10 mm,
    dashed,
    node contents={}}
}

\begin{tikzpicture}[node distance=2cm, auto, thick, scale = 0.5]

\begin{scope}
\node (I1) [draw, semicircle, rotate = 90] {\rotatebox{270}{$\mathsf{I}_{t-1}$} };
\node (i1) [draw, red, semicircle, rotate = 270] at ($(I1)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{t-1}$} };
\node (I2) [draw, semicircle, right of = i1, rotate = 90] {\rotatebox{270}{$\mathsf{I}_{t}(\mathsf{i}_{t-1})$} };
\end{scope}

\begin{scope}
\node (R1) [draw, right of = I2, semicircle, rotate = 90] {\rotatebox{270}{$R_{t - 1}$} };
\node (r1) [draw, red, semicircle, rotate = 270] at ($(R1)+(60pt,0)$) {\rotatebox{90}{$r_{t-1}$} };
\node (C1) [draw, semicircle, rotate = 90] at ($(R1)+(0pt, 150pt)$) {\rotatebox{270}{$C_{t - 1}$} };
\node (c1) [draw, red, semicircle, rotate = 270] at ($(C1)+(60pt,0)$) {\rotatebox{90}{$c_{t-1}$} };
\node (P1) [draw, semicircle, rotate = 90] at ($(R1)+(0pt,-150pt)$) {\rotatebox{270}{$P_{t - 1}$} };
\node (p1) [draw, red, semicircle, rotate = 270] at ($(P1)+(60pt,0)$) {\rotatebox{90}{$p_{t-1}$} };
\node (C2) [draw, ellipse, rotate = 90] at ($(c1)+(300pt,0)$) {\rotatebox{270}{$C_{t}(c_{t-1}, r_{t-1}, p_{t-1})$} };
\node (R2) [draw, ellipse, rotate = 90] at ($(r1)+(300pt,0)$) {\rotatebox{270}{$R_{t}(c_{t-1}, r_{t-1}, p_{t-1})$} };
\node (P2) [draw, ellipse, rotate = 90] at ($(p1)+(300pt,0)$) {\rotatebox{270}{$P_{t}(c_{t-1}, r_{t-1}, p_{t-1})$} };
\end{scope}

\path[->, blue] (i1) edge (I2);

\path[->, blue] (c1) edge (C2);
\path[->, blue] (c1) edge (R2);
\path[->, blue] (c1) edge (P2);
\path[->, blue] (r1) edge (C2);
\path[->, blue] (r1) edge (R2);
\path[->, blue] (r1) edge (P2);
\path[->, blue] (p1) edge (C2);
\path[->, blue] (p1) edge (R2);
\path[->, blue] (p1) edge (P2);

\end{tikzpicture}
```


```{r tikz-swit, eval = TRUE, engine = "tikz", fig.cap = "Intervention template for a single plant evolving over time. Time two timepoints are shown. The plant has two leaves in the first week, and third leaf emerges in the second week.", fig.ext = 'pdf', echo = FALSE}
\usetikzlibrary{arrows, shapes, calc, shapes.geometric, fit}

\tikzset{
container/.style = {
    rectangle,
    draw,
    inner sep = 2 mm,
    dashed,
    node contents={}}
}

\begin{tikzpicture}[node distance=2cm, auto, thick, scale = 0.5]

\begin{scope}
\node (I11) [draw, semicircle, rotate = 90] {\rotatebox{270}{$\mathsf{I}_{11}$} };
\node (i11) [draw, red, semicircle, rotate = 270] at ($(I11)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{11}$} };
\node (I21) [draw, semicircle, rotate = 90] at ($(I11)+(0pt,-100pt)$) {\rotatebox{270}{$\mathsf{I}_{21}$} };
\node (i21) [draw, red, semicircle, rotate = 270] at ($(I21)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{21}$} };
\node (L1)  [draw, circle] at ($(I11)+(-40pt,70pt)$) {$L_1$};

\node (I12) [draw, right of = i11, semicircle, rotate = 90] {\rotatebox{270}{$\mathsf{I}_{12}(i_{1})$} };
\node (i12) [draw, red, semicircle, rotate = 270] at ($(I12)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{12}$} };
\node (I22) [draw, right of = i21, semicircle, rotate = 90] {\rotatebox{270}{$\mathsf{I}_{22}(\mathsf{i}_{1})$} };
\node (i22) [draw, red, semicircle, rotate = 270] at ($(I22)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{22}$} };
\node (I32) [draw, semicircle, rotate = 90] at ($(I22)+(0pt,-100pt)$) {\rotatebox{270}{$\mathsf{I}_{32}(\mathsf{i}_{1})$} };
\node (i32) [draw, red, semicircle, rotate = 270] at ($(I32)+(60pt,0)$) {\rotatebox{90}{$\mathsf{i}_{32}$} };
\end{scope}

\path[->, blue] (i11) edge (I12);
\path[->, blue] (i11) edge (I22);
\path[->, blue] (i11) edge (I32);
\path[->, blue] (i21) edge (I12);
\path[->, blue] (i21) edge (I22);
\path[->, blue] (i21) edge (I32);

\path[->, black] (L1) edge (I11);
\path[->, black, bend left] (L1) edge (I12);
\path[->, black] (L1) edge (I21);
\path[->, black, bend left] (L1) edge (I22);
\path[->, black, bend right] (L1) edge (I32);

\node [container,
       fit=(L1) (i32),
       label=left: A plant];


\end{tikzpicture}
```

## Assumptions of this causal model

* no carryover effects (Markov assumption)

## Experiment

Describe the experiment that was run:

* experimental treatments
* what was measured

Describe study predictions or hypotheses.

## Estimation

Exposure model: 

* good case for mixed-effects multivariate Bernoulli model [@dai2013multivariate]? i.e., model the probability of an individual leaf's infection statuses as:

\[
h\{\Pr(C_{jkt} = c, R_{jkt}  = r, P_{jkt}  = p | L_{jkt}, b_{jt})|\}
\]

where $h$ is (e.g.) the inverse logistic function. Then maybe estimate the plant's infection status as:

\[
\Pr(\mathsf{I}_{jt} = \mathsf{i}_{jt} | L_{jt}) = \int \prod_{k = 1}^{n_j}h\{\Pr(\mathsf{I}_{jkt} = \mathsf{i}_{jkt}  | L_{jkt}, b_{jt})\}^{I(\mathsf{I}_{jkt} = \mathsf{i}_{jkt} )} f(b_{jt}) \mathsf{d}b_{jt}
\]


* Description of statistical models
  * @barkley2017causal
  * account for ordering of leaves within host
* M-estimation for variance estimates

# Results

* Sensitively + stability analyses

# Discussion

* Discussion on mechanisms:
  * Data from this experiment cannot distinguish between niche preemption or niche modification. Needs additional assumptions and knowledge, and this is even more complicated if a species can change from (e.g.) biotroph to necrotroph.
  * mediation - how an effect occurs; interaction - for whom an effect occurs [@vanderweele2015explanation]
* suggestions for better designs and how identifiability results might suggest better design
* links to other causal frameworks
* link back to @halliday2017interactions
  * is death of leaf a censoring event or competing risk? There's a complication that leaves only live a certain period of time - during which they are either infected or not - but it is not as if a leaf *could* survive for longer during which time we could observe the outcome. Then there's the fact that leaves in the last cohort can't be observed for the same length of time as other cohorts due to end of study and senesence. 
* Structural equation models [@shipley2016cause] are popular in biology and ecology; however, parameters from these models may not carry a causal interpretation without assumptions often implicitly made [@imai2010general]. Similarly, it is not uncommon for authors to discuss the results of associational statistical models in terms of causal language. 


   
# References