---
title: "Playing around with multistate models"
author: "Bradley Saul"
date: "9/21/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(msm)
dt <- readRDS("../data/analysis_data.rds")
# dt <- readRDS("data/analysis_data.rds")
```


```{r, echo = FALSE}
set_remaining_infected <- function(x){
  # finds the first element of x == 1, then sets any elements after that to 1
  n <- length(x)
  if(sum(x) == 0 || (sum(x) == 1 & x[n] == 1)){ return(x) }
  
  # Find first x[i] == 1
  s <- min(which(x == 1))
  
  x[(s+1):n] <- 1
  x
}

```

```{r, echo = FALSE}
dt %>%
  group_by(LeafID) %>%
  # Remove any with a single obsevation
  filter(n() > 1) %>%
  select(-leaf_damage) %>%
  group_by(ID, Tiller) %>%
  tidyr::spread(key = pathogen, value = infected) %>%
  mutate(
    t        = days_since_trial_start,
    leaf_age = days_since_emergence,
    w        = as.integer(weeks_since_trial_start)
    # leaf_age = as.integer(weeks_since_emergence)
  ) %>%
  select(LeafID, ID, Tiller, Leaf, t, leaf_age, cohort, current_experimental_period,
         C = Col, R = Rhiz, P = Rust)  %>% 
  group_by(ID, Tiller, Leaf) %>% 
  arrange(t) %>%
  mutate(
    emergent_experimental_period = current_experimental_period[1],
    C = set_remaining_infected(C),
    R = set_remaining_infected(R),
    P = set_remaining_infected(P)
  ) %>% 
  group_by(ID, Tiller, t) %>%
  # Compute proportion of other leaves on same tiller infected with C & R
  # at time t
  # TODO: this needs to be changed to find proportion of leaves in past X number of days.
  mutate(
    D  = 0L,
    n_leaves = n(),
    pC = if_else(n_leaves > 1, (sum(C) - C)/(n_leaves - 1), 0),
    pR = if_else(n_leaves > 1, (sum(R) - R)/(n_leaves - 1), 0),
    pP = if_else(n_leaves > 1, (sum(P) - P)/(n_leaves - 1), 0),
    sC = case_when(
      0   <= pC & pC <= 0.25 ~ 0,
      0.25 <= pC & pC <= 1   ~ 1),
    sR = case_when(
      0   <=  pR & pR <= 0.25 ~ 0,
      0.25 <= pR & pR <= 1   ~ 1),
    sP = case_when(
      0   <=  pP & pP <= 0.25 ~ 0,
      0.25 <= pP & pP <= 1   ~ 1)
  ) %>% 
  split(., .$LeafID) %>%
  purrr::map_dfr(
    .f = function(dt){
      add_row(
        ungroup(dt),
        LeafID = dt$LeafID[1],
        ID     = dt$ID[1],
        Tiller = dt$Tiller[1],
        Leaf   = dt$Leaf[1],
        emergent_experimental_period = dt$emergent_experimental_period[1],
        current_experimental_period = dt$current_experimental_period[1],
        t      = max(dt$t) + 1,
        leaf_age = max(dt$leaf_age) + 1,
        cohort = dt$cohort[1],
        D      = 1L
      )
    }
  ) -> hold

```


# Multistate models

Let $(C, C^{\star}, R)$ be the infection state of a leaf where $C^{\star}$ is the an indicator that the proportion of other leaves on the same plant infected with $C$ is $>0.25$.

Consider models for the following two settings, one including $C^{\star}$ (simple), another including $C^{\star}$ (interference).

The simple model has 5 states:

* 1 = (C = 0, R = 0)
* 2 = (C = 1, R = 0)
* 3 = (0, 1)
* 4 = (1, 1)
* 5 = X - death


The inteference model has 9 states:

* 1 = $(C = 0, C^{\star} = 0, R = 0)$
* 2 = (1, 0, 0)
* 3 = (0, 1, 0)
* 4 = (0, 0, 1)
* 5 = (1, 1, 0)
* 6 = (0, 1, 1)
* 7 = (1, 0, 1)
* 8 = (1, 1, 1)
* 9 = X - death


Assumptions

* no recovery (leaf can't become uninfected by a pathogen)


```{r, echo = TRUE}
analysis_dt <- hold %>%
  mutate(
    state_simple    = if_else(D == 1, "X", paste(C, R, sep = "-")),
    state_simpleL = case_when(
      state_simple == "0-0" ~ 1L,
      state_simple == "1-0" ~ 2L,
      state_simple == "0-1" ~ 3L,
      state_simple == "1-1" ~ 4L,
      state_simple == "X"   ~ 5L
    ),
    state_interfere = if_else(D == 1, "X", paste(C, sC, R, sep = "-")),
    state_interfereL = case_when(
      state_interfere == "0-0-0" ~ 1L,
      state_interfere == "1-0-0" ~ 2L,
      state_interfere == "0-1-0" ~ 3L,
      state_interfere == "0-0-1" ~ 4L,
      state_interfere == "1-1-0" ~ 5L,
      state_interfere == "0-1-1" ~ 6L,
      state_interfere == "1-0-1" ~ 7L,
      state_interfere == "1-1-1" ~ 8L,
      state_interfere == "X"     ~ 9L
    )
  )
```



```{r, echo = FALSE}

Q_simple <- rbind(
  c(0, .1, .1, .1, .1),
  c(0,  0,  0,  .1, .1),
  c(0,  0,  0,  .1, .1),
  c(0,  0,  0,  .1, .1),
  c(0,  0,  0,   0, 0)
)

Q_interfere <- rbind(
  c(0, .1, .1, .1,  0,  0,  0,  0, .1),
  c(0,  0,  0,  0, .2,  0, .1, .1, .1),
  c(.1,  .1,  0,  0, .2, .1,  0, .1, .1),
  c(0,  0,  0,  0,  0, .1, .1, .1, .1),
  c(0,  .1,  0,  0,  0,  0,  0, .1, .1),
  c(0,  0,  0,  .1,  0,  0,  0, .1, .1),
  c(0,  0,  0,  0,  0,  0,  0, .1, .1),
  c(0,  0,  0,  0,  0,  0,  .1, 0, .1),
  c(0,  0,  0,  0,  0,  0,  0, 0, 0)
)
```

```{r, echo = FALSE}
fit_msm <- function(formula, subject, data, covariates = NULL, Q){
  Q_crude <- crudeinits.msm(
    formula = formula,
    subject = subject,
    data    = data,
    qmatrix = Q)

  fit <- msm(
    formula    = formula,
    subject    = subject,
    covariates = covariates, 
    data       = data,
    qmatrix    = Q_crude)
  
  fit
}
```



```{r}
simple_msm <- fit_msm(
  formula  = state_simpleL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  Q        = Q_simple
)


interfere_msm <- fit_msm(
  formula  = state_interfereL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  Q        = Q_interfere
)

censor_msm <- msm(
  formula  = state_simpleL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  qmatrix    = Q_simple,
  censor = 5)
```

```{r}
compare_probs <- function(msmfit, t, i1, i2, label){
  n  <- length(t)
  P  <- pmatrix.msm(msmfit, t = t)
  p1 <- apply(P, 3, function(x) x[i1[1], i1[2]])
  p2 <- apply(P, 3, function(x) x[i2[1], i2[2]])
  
  data_frame(
    label = label,
    comparison = sprintf("Pr(%s|%s)vPr(%s|%s)", i1[2], i1[1], i2[2], i2[1]),
    t = t,
    p1 = p1,
    p2 = p2,
    rd = p1 - p2,
    rr = p1/p2
  )
}
```


```{r}

X <- bind_rows(
    compare_probs(censor_msm, t = 1:35, i1 = c(3,4), i2 = c(3,3),
                   label = "Censor on Death: From (!C,R) to (C,R) given (!C, R)"),
  compare_probs(simple_msm, t = 1:35, i1 = c(3,4), i2 = c(3,3),
                   label = "From (!C,R) to (C,R) given (!C, R)"),
  compare_probs(interfere_msm, t = 1:35, i1 = c(4,7), i2 = c(4,4),
                   label = "From (!C,!C*,R) to (C,!C*,R) given (!C,!C*,R)"),
  compare_probs(interfere_msm, t = 1:35, i1 = c(6,8), i2 = c(6,6),
                 label = "From (!C,C*,R) to (C,C*,R) given (!C,C*,R)")
)

X %>% 
  tidyr::gather(key = "metric", value = "value", -label, -comparison, -t) %>%
  ggplot(
    data = .,
    aes(x = t, y = value, group = label, color = label)
  ) + geom_line() + 
    facet_grid(metric ~ ., scales = "free_y")


```


```{r}
pnext.msm(simple_msm)[3,4]
pnext.msm(interfere_msm, ci = "delta")[4, 7]
pnext.msm(interfere_msm, ci = "delta")[6, 8]  
   
```

```{r eval = FALSE}
vals <- function(t){
  PP_simple    <- pmatrix.msm(msm_simple,    t = t) 
                              # covariates = list(emergent_experimental_period = 2))
  PP_interfere <- pmatrix.msm(msm_interfere, t = t)
                              # covariates = list(emergent_experimental_period = 2))
  
  P11gP01   <- PP_simple[3, 4]
  P01gP01   <- PP_simple[3, 3]
  P101gP001 <- PP_interfere[4 , 7] 
  P001gP001 <- PP_interfere[4, 4]
  P111gP011 <- PP_interfere[6 , 8]
  P011gP011 <- PP_interfere[6, 6]
  
  data_frame(
    t         = t,
    metric    = c("p1", "p2", "rd", "rr"),
    P11vP01   = c(P11gP01, P01gP01, P11gP01 - P01gP01, P11gP01/P01gP01),
    P101vP001 = c(P101gP001, P001gP001, P101gP001 - P001gP001, P101gP001/P001gP001),
    P111vP011 = c(P111gP011, P011gP011, P111gP011 - P011gP011, P111gP011/P011gP011)
  )
}

plot_dt <- purrr::map_dfr(1:21, ~ vals(.x)) %>%
  tidyr::gather(key = "trans", value = "val", -t, -metric)


ggplot(
  data = plot_dt,
  aes(x = t, y = val, group = trans, color = trans)
) + geom_line() + 
  facet_grid(metric ~ ., scales = "free_y")

```
