---
title: "Playing around with multistate models"
author: "Bradley Saul"
date: "9/21/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(msm)
dt <- readRDS("../data/analysis_data.rds")
# dt <- readRDS("data/analysis_data.rds")
```


```{r, echo = FALSE}
set_remaining_infected <- function(x){
  # finds the first element of x == 1, then sets any elements after that to 1
  n <- length(x)
  if(sum(x) == 0 || (sum(x) == 1 & x[n] == 1)){ return(x) }
  
  # Find first x[i] == 1
  s <- min(which(x == 1))
  
  x[(s+1):n] <- 1
  x
}

```

```{r, echo = FALSE}
dt %>%
  group_by(LeafID) %>%
  # Remove any with a single obsevation
  filter(n() > 1) %>%
  select(-leaf_damage) %>%
  group_by(ID, Tiller) %>%
  tidyr::spread(key = pathogen, value = infected) %>%
  mutate(
    t        = days_since_trial_start,
    leaf_age = days_since_emergence,
    w        = as.integer(weeks_since_trial_start)
    # leaf_age = as.integer(weeks_since_emergence)
  ) %>%
  select(LeafID, ID, Tiller, Leaf, t, leaf_age, cohort, current_experimental_period,
         C = Col, R = Rhiz, P = Rust)  %>% 
  group_by(ID, Tiller, Leaf) %>% 
  arrange(t) %>%
  mutate(
    emergent_experimental_period = current_experimental_period[1],
    C = set_remaining_infected(C),
    R = set_remaining_infected(R),
    P = set_remaining_infected(P)
  ) %>% 
  group_by(ID, Tiller, t) %>%
  # Compute proportion of other leaves on same tiller infected with C & R
  # at time t
  # TODO: this needs to be changed to find proportion of leaves in past X number of days.
  mutate(
    D  = 0L,
    n_leaves = n(),
    pC = if_else(n_leaves > 1, (sum(C) - C)/(n_leaves - 1), 0),
    pR = if_else(n_leaves > 1, (sum(R) - R)/(n_leaves - 1), 0),
    pP = if_else(n_leaves > 1, (sum(P) - P)/(n_leaves - 1), 0),
    sC = case_when(
      0   <= pC & pC <= 0.25 ~ 0,
      0.25 <= pC & pC <= 1   ~ 1),
    sR = case_when(
      0   <=  pR & pR <= 0.25 ~ 0,
      0.25 <= pR & pR <= 1   ~ 1),
    sP = case_when(
      0   <=  pP & pP <= 0.25 ~ 0,
      0.25 <= pP & pP <= 1   ~ 1)
  ) %>% 
  split(., .$LeafID) %>%
  purrr::map_dfr(
    .f = function(dt){
      add_row(
        ungroup(dt),
        LeafID = dt$LeafID[1],
        ID     = dt$ID[1],
        Tiller = dt$Tiller[1],
        Leaf   = dt$Leaf[1],
        emergent_experimental_period = dt$emergent_experimental_period[1],
        current_experimental_period = dt$current_experimental_period[1],
        t      = max(dt$t) + 1,
        leaf_age = max(dt$leaf_age) + 1,
        cohort = dt$cohort[1],
        D      = 1L
      )
    }
  ) -> hold

```


# Multistate models

Let $(C, C^{\star}, R)$ be the infection state of a leaf where $C^{\star}$ is the an indicator that the proportion of other leaves on the same plant infected with $C$ is $>0.25$.

Consider models for the following two settings, one including $C^{\star}$ (simple), another including $C^{\star}$ (interference).

The simple model has 5 states:

* 1 = (C = 0, R = 0)
* 2 = (C = 1, R = 0)
* 3 = (0, 1)
* 4 = (1, 1)
* 5 = X - death


The inteference model has 9 states:

* 1 = $(C = 0, C^{\star} = 0, R = 0)$
* 2 = (1, 0, 0)
* 3 = (0, 1, 0)
* 4 = (0, 0, 1)
* 5 = (1, 1, 0)
* 6 = (0, 1, 1)
* 7 = (1, 0, 1)
* 8 = (1, 1, 1)
* 9 = X - death


Assumptions

* no recovery (leaf can't become uninfected by a pathogen)


```{r, echo = TRUE}
analysis_dt <- hold %>%
  mutate(
    state_simple    = if_else(D == 1, "X", paste(C, R, sep = "-")),
    state_simpleL = case_when(
      state_simple == "0-0" ~ 1L,
      state_simple == "1-0" ~ 2L,
      state_simple == "0-1" ~ 3L,
      state_simple == "1-1" ~ 4L,
      state_simple == "X"   ~ 5L
    ),
    state_censorL = if_else(state_simpleL == 5L, 999L, state_simpleL),
    state_interfere = if_else(D == 1, "X", paste(C, sC, R, sep = "-")),
    state_interfereL = case_when(
      state_interfere == "0-0-0" ~ 1L,
      state_interfere == "1-0-0" ~ 2L,
      state_interfere == "0-1-0" ~ 3L,
      state_interfere == "0-0-1" ~ 4L,
      state_interfere == "1-1-0" ~ 5L,
      state_interfere == "0-1-1" ~ 6L,
      state_interfere == "1-0-1" ~ 7L,
      state_interfere == "1-1-1" ~ 8L,
      state_interfere == "X"     ~ 9L
    )
  )
```


The following are the allowed transitions for each model. Zeros indicate a transition from state $i$ to $j$ is not permitted (except for diagonals -- leaves remain in the same state). The non-zero values don't matter.

```{r, echo = TRUE}
Q_censor <- rbind(
  c(0, 1, 1, 1),
  c(0, 0, 1, 1),
  c(0, 0, 0, 1),
  c(0, 0, 0, 0)
)

Q_simple <- rbind(
  c(0, 1,  1, 1, 1),
  c(0, 0,  0, 1, 1),
  c(0, 0,  0, 1, 1),
  c(0, 0,  0, 1, 1),
  c(0, 0,  0, 0, 0)
)

Q_interfere <- rbind(
  c(0, 1, .1, .1,  0,  0,  0,  0, .1),
  c(0,  0,  0,  0, .2,  0, .1, .1, .1),
  c(.1,  .1,  0,  0, .2, .1,  0, .1, .1),
  c(0,  0,  0,  0,  0, .1, .1, .1, .1),
  c(0,  .1,  0,  0,  0,  0,  0, .1, .1),
  c(0,  0,  0,  .1,  0,  0,  0, .1, .1),
  c(0,  0,  0,  0,  0,  0,  0, .1, .1),
  c(0,  0,  0,  0,  0,  0,  .1, 0, .1),
  c(0,  0,  0,  0,  0,  0,  0, 0, 0)
)
```

```{r, echo = FALSE}
fit_msm <- function(formula, subject, data, covariates = NULL, 
                    startQ = NULL, Q = NULL,
                    censor = NULL){
  
  if(is.null(Q)){
    Q_crude <- crudeinits.msm(
        formula = formula,
        subject = subject,
        data    = data,
        censor  = censor,
        qmatrix = startQ)
  } else {
    Q_crude = Q
  }

  
  fit <- msm(
    formula    = formula,
    subject    = subject,
    covariates = covariates, 
    data       = data,
    censor     = censor,
    qmatrix    = Q_crude)
  
  fit
}
```



```{r}
simple_msm <- fit_msm(
  formula  = state_simpleL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  startQ   = Q_simple
)

interfere_msm <- fit_msm(
  formula  = state_interfereL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  startQ   = Q_interfere
)

Qcen <- crudeinits.msm(
  formula  = state_censorL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  censor   = 999,
  qmatrix  = Q_censor)

# Hack to get optimization to converge
Qcen[4, 4] <- -.01
Qcen[4, 3] <- .01

censor_msm <- fit_msm(
  formula  = state_censorL ~ leaf_age,
  subject  = analysis_dt$LeafID,
  data     = analysis_dt,
  Q        = Qcen,
  censor   = 999)
```

```{r, echo = FALSE}
compare_probs <- function(msmfit, t, i1, i2, label){
  n  <- length(t)
  P  <- pmatrix.msm(msmfit, t = t)
  p1 <- apply(P, 3, function(x) x[i1[1], i1[2]])
  p2 <- apply(P, 3, function(x) x[i2[1], i2[2]])
  
  data_frame(
    label = label,
    comparison = sprintf("Pr(%s|%s)vPr(%s|%s)", i1[2], i1[1], i2[2], i2[1]),
    t = t,
    p1 = p1,
    p2 = p2,
    rd = p1 - p2,
    rr = p1/p2
  )
}
```

In the following plot, `p1` is $\Pr(\text{in state } j \text{ by time } t | \text{in state } i \text{ now}  )$. `p2` is $\Pr(\text{in state } i \text{ by time } t | \text{in state } i \text{ now}  )$ (i.e., the probability of remaining in the same state). `rd` is the risk difference `p1 - p2` and `rr` is the risk ratio `p1/p2`. The model is a continuous time markov model so the interpretation of the transition probabilities are: given that a host in state $i$ at time $t^{\star}$, what is the probability of transition to state $j$ at time $t^{\star} + t$.

In the figure, I compare the following transition probabilities:

* becoming infected with $C$ given that a host is current not infected by $C$ but is infected by $R$ (state 3 $\to$ 4 in the simple model and censoring model)
* becoming infected with $C$ given that a host is current not infected by $C$ but is infected by $R$ *and is not* surrounded by other $C$ infected leaves (state 4 $\to$ 6 in the interference model)
* becoming infected with $C$ given that a host is current not infected by $C$ but is infected by $R$ *and is* surrounded by other $C$ infected leaves (state 6 $\to$ 8 in the interference model)

to the probability of staying in the first state (e.g. remaining in state 3 for the simple model).

```{r, echo = FALSE, fig.width=7}
X <- bind_rows(
  compare_probs(censor_msm, t = 1:35, i1 = c(3,4), i2 = c(3,3),
                   label = "From 3 to 4 given 3 (Censored on Death)"),
  compare_probs(simple_msm, t = 1:35, i1 = c(3,4), i2 = c(3,3),
                   label = "From 3 to 4 given 4"),
  compare_probs(interfere_msm, t = 1:35, i1 = c(4,7), i2 = c(4,4),
                   label = "From 4 to 7 given 4"),
  compare_probs(interfere_msm, t = 1:35, i1 = c(6,8), i2 = c(6,6),
                 label = "From 6 to 8 given 6")
)

X %>% 
  tidyr::gather(key = "metric", value = "value", -label, -comparison, -t) %>%
  ggplot(
    data = .,
    aes(x = t, y = value, group = label, color = label)
  ) + geom_line() + 
    facet_grid(metric ~ ., scales = "free_y")
```

The results suggest a few things:

* treating death as a censoring event targets an estimand that is probably not relevant (it assumes infection could be observed at some point after death)
* what is happening on other leaves within the same plant is relevant (i.e. interference does appear to be at play here)
* the definition of the "priority effect" needs consideration of the scale of the contrast. The probability of these states is shrinkingly small as a leaf ages, so is a relative risk of two unlikely events the relevant measure for a priority effect?


```{r, eval = FALSE, echo = FALSE}
pnext.msm(censor_msm)[3,4]
pnext.msm(simple_msm)[3,4]
pnext.msm(interfere_msm, ci = "delta")[4, 7]
pnext.msm(interfere_msm, ci = "delta")[6, 8]  
```
